---
title: 'Prophet (time series)'
author: 'Charlie Kopp'
date: today
date-format: long
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false

---

```{r}
#| label: setup
#| include: false

# install.packages(c("prophet", "data.table"))
library(data.table)
library(arrow)
dir.create("reports/figures/forecasting", recursive = TRUE, showWarnings = FALSE)
```

# Data

```{r}
#| label: read data

dt <- read_parquet('data/processed/dt_clean.parquet')
user_outcomes <- read_parquet('data/processed/user_outcomes.parquet')
```

***Thomas Feedback:***

- predicting proportion of success based on first action date is too complicated. just make outcome the total orders shipped on any given day.
- also need to figure out how prophet handles missing data (no orders are shipped on weekends)

There's a few **issues**:

- the first couple months have no orders_shipped: ongoing journeys weren't included in dataset. for accurate time series need to figure out start point where most journeys that were ongoing would have wrapped up. 
  - solution: take 80th pctl of successful journey length

- there should be no data on weekends (warehouse closed = no order_shipped). but there are actually low counts on some weekends, missing on others

## Total orders shipped per-day approach

```{r}
#| label: format data

# just want to summarize data for total orders shipped per-day

prophet_train_data <- dt[, .(
  y = sum(event_name=="order_shipped")
), by = .(ds = as.IDate(event_timestamp))]

new_start_date <- min(prophet_train_data$ds) + 71 # = 80th pctl of j-length, found below

prophet_train_data <- prophet_train_data[ds > new_start_date]
setorder(prophet_train_data, ds)
```

### Find start date where ongoing orders would have shipped by then

- 80th percentile of successful journey length = 71 days

```{r}
# label: figure out start cutoff

# 1. solve start cutoff issue

success_id <- user_outcomes[final_outcome == "success", id]

success_lengths <- dt[id %in% success_id,{
  first_date <- as.IDate(event_timestamp[1])
	last_date <- as.IDate(event_timestamp[.N])
  .(
    journey_length_days = as.integer(last_date - first_date, units='days')
  )
}, by=id]

hist(success_lengths$journey_length_days, breaks = 100, freq=FALSE)

# find the 95th percentile of how long it takes people to buy
quantile(success_lengths$journey_length_days, probs = 0.95, na.rm = TRUE) # 95th pctile = 221 days
quantile(success_lengths$journey_length_days, probs = 0.80, na.rm = TRUE) # 80th pctile = 71 days

new_start_date <- min(prophet_train_data$ds) + 71

prophet_train_data <- prophet_train_data[ds > new_start_date]
```

### deal with weekend entries

```{r}
# label: fix weekend counts

weekend_dat <- prophet_train_data[weekdays(ds) == "Saturday" | weekdays(ds) == "Sunday"]

plot(weekend_dat$ds, weekend_dat$y)

setorder(weekend_dat, -y) |> head(20)

summary.default(weekend_dat$y)

# easy solution: drop weekends

daily_train_data <- prophet_train_data[!(ds %in% weekend_dat$ds)]
```

Most of the weekend data is low (~ 56), but some are high (mostly in later months).

## Total orders shipped per-WEEK

```{r}
#| label: weekly data

weekly_train_data <- prophet_train_data[,
  .(y = sum(y)),
  by = .(ds = round(ds, "weeks"))
]
```

## Total orders shipped per-MONTH


```{r}
#| label: monthly orders data

monthly_train_data <- prophet_train_data[, .(y = sum(y)), by = .(ds = round(ds, "months"))]

# cut incomplete first and last months
monthly_train_data <- monthly_train_data[ds > "2021-01-01" & ds < "2023-01-01"]
```

## First-action date p-success approach

```{r}
#| label: format data (old)

# drop incomplete rows

train_cutoff_date <- as.IDate(max(dt$event_timestamp))

# flatten data by user

user_dt <- dt[, .(
    first_action_date = as.IDate(min(event_timestamp)),
    last_action_date = as.IDate(max(event_timestamp)),
    has_shipped = any(event_name == "order_shipped"),
    total_actions = .N
), by = id]

# define days since last action as days_inactive

user_dt[, days_inactive := as.integer(train_cutoff_date - last_action_date)]

user_dt[, journey_outcome := fcase(
    has_shipped == TRUE, "success",
    has_shipped == FALSE & days_inactive > 60, "failure",
    default = "incomplete"
)]

table(user_dt$journey_outcome)

# drop incomplete journeys

user_dt_complete <- user_dt[journey_outcome != "incomplete", ]

# want to find distr. journey length for a successes and add upper end to cutoff date to reduce survivorship bias

successes_only <- user_dt_complete[journey_outcome == "success"]

purchase_times <- dt[event_name == "order_shipped", 
                     .(purchase_date = as.IDate(min(event_timestamp))), 
                     by = id]

# join back to find the journey length
journey_lengths <- merge(successes_only, purchase_times, by = "id")
journey_lengths[, days_to_buy := as.integer(purchase_date - first_action_date)]

hist(journey_lengths$days_to_buy, breaks = 100, freq=FALSE)

# find the 95th percentile of how long it takes people to buy
p95_days_success <- quantile(journey_lengths$days_to_buy, probs = 0.95, na.rm = TRUE)
print(paste("95% of successful users buy within", round(p95_days_success), "days"))

p80_days_success <- quantile(journey_lengths$days_to_buy, probs = 0.80, na.rm = TRUE)

# define safe cohort start date
# Safe Date = Cutoff - (95th Percentile Days + 60 Days Inactivity)
safe_cutoff <- train_cutoff_date - (round(p95_days_success) + 60)

cohort_users <- user_dt_complete[first_action_date <= safe_cutoff]

prophet_train_data <- cohort_users[, .(
    total_users = .N,
    successes = sum(journey_outcome == "success"),
    y = sum(journey_outcome == "success") / .N # Prophet requires the target to be named 'y'
), by = .(ds = first_action_date)] # Prophet requires the date to be named 'ds'
```

# Model fitting

```{r}
#| label: load prophet package

library(prophet)
```

## fit orders-per-DAY model

```{r}
#| label: fit prophet model to daily data

# sort data by time
setorder(daily_train_data, ds)

# Initialize Prophet for daily data
m <- prophet(
  yearly.seasonality = TRUE,
  weekly.seasonality = TRUE,
  daily.seasonality = FALSE
)

m <- add_country_holidays(m, country_name = "US")

m <- fit.prophet(m, daily_train_data)

# Calculate DAYS needed to reach Dec 2024
last_ds <- max(daily_train_data$ds)
n_days <- length(seq(last_ds, as.IDate("2024-12-01"), by = "day")) - 1

# Create future dataframe at daily frequency
future_daily <- make_future_dataframe(m, periods = max(0, n_days), freq = "day")
forecast_daily <- predict(m, future_daily)
```

## fit orders-per-WEEK model

```{r}
#| label: prophet weekly model

# sort data by time
setorder(weekly_train_data, ds)

# Initialize Prophet for weekly data
m <- prophet(
  yearly.seasonality = TRUE,
  weekly.seasonality = FALSE,
  daily.seasonality = FALSE
)

m <- fit.prophet(m, weekly_train_data)

# Calculate WEEKS needed to reach Dec 2024
last_ds <- max(weekly_train_data$ds)
n_weeks <- length(seq(last_ds, as.IDate("2024-12-01"), by = "week")) - 1

# Create future dataframe at daily frequency
future_weekly <- make_future_dataframe(m, periods = max(0, n_weeks), freq = "week")
forecast_weekly <- predict(m, future_weekly)
```

## fitting to orders-per-MONTH
```{r}
#| label: fit prophet model to monthly data

# 1. bind target to prevent negative predictions
monthly_train_data <- monthly_train_data %>%
  mutate(
    cap = max(y) * 1.5, # Set a reasonable upper bound for the target
    floor = 0
  )

# 2. Initialize the model with DEFAULT yearly seasonality turned OFF
m_monthly <- prophet(
  growth = 'logistic',   # Turn on bounded growth
  yearly.seasonality = FALSE, 
  weekly.seasonality = FALSE, 
  daily.seasonality = FALSE,
  changepoint.prior.scale = 0.01 # Makes the trend line much stiffer
)

# 3. Manually add a "gentle" yearly seasonality
# Setting fourier.order to 2 or 3 forces a smooth, gentle seasonal wave
# rather than the jagged, over-fitted 10-order default.
m_monthly <- add_seasonality(m_monthly, name = 'yearly', period = 365.25, fourier.order = 2)

# 4. Fit the model
m_monthly <- fit.prophet(m_monthly, monthly_train_data)

# 5. Predict
future_monthly <- make_future_dataframe(m_monthly, periods = 24, freq = 'month')
future_monthly <- future_monthly %>%
  mutate(
    cap = max(monthly_train_data$y) * 1.5, # Use the same cap for future data
    floor = 0
  )
forecast_monthly <- predict(m_monthly, future_monthly)

```

## fitting to cohort method

```{r}
#| label: fit prophet model

m <- prophet(yearly.seasonality = TRUE, weekly.seasonality=TRUE)
m <- add_country_holidays(m, country_name = "US")
m <- fit.prophet(m, prophet_train_data)

# calculate exactly how many days of "incomplete" cohorts we need to forecast
days_to_forecast <- as.numeric(train_cutoff_date - safe_cutoff)

# create a dataframe containing both historical dates AND future dates
future_dates <- make_future_dataframe(m, periods = days_to_forecast)

# generate the forecast
# This creates a massive dataframe containing the prediction (yhat) and all components
forecast <- predict(m, future_dates)
```

# Interpretation / Visualization

```{r}
#| label: load viz packages

library(ggplot2)
library(dplyr)
library(patchwork)
```

## Daily orders shipped 

```{r}
#| label: plot daily model

# plot(prophet_train_data$ds, prophet_train_data$y, cex=0.5)

# sum(weekdays(prophet_train_data$ds) %in% c("Saturday", "Sunday"))

# setorder(prophet_train_data, y) |> head(20)

plot(m, forecast_daily) +
  labs(title = "Daily Orders Forecast through 2024", x = "Date", y = "Orders Shipped") +
  theme_minimal()

ggsave("reports/figures/forecasting/daily_forecast.png", width=8, height=4)

# Plot 2: The Components Breakdown
comp_plots <- prophet_plot_components(m, forecast_daily)

wrap_plots(comp_plots, ncol=1)

ggsave("reports/figures/forecasting/daily_components.png", width=8, height=6)
```

## Weekly orders shipped

```{r}
#| label: plot weekly model

plot(m, forecast_weekly) +
  labs(title = "Daily Orders Forecast through 2024", x = "Date", y = "Orders Shipped") +
  theme_minimal()

ggsave("reports/figures/forecasting/weekly_forecast.png", width=8, height=4)

# Plot 2: The Components Breakdown
prophet_plot_components(m, forecast_weekly)
```

## Monthly orders shipped 

```{r}
#| label: plot monthly model


# Visualize monthly forecast
plot(m_monthly, forecast_monthly) +
  labs(title = "Monthly Orders Forecast through 2024", x = "Month", y = "Orders Shipped") +
  theme_minimal()

ggsave("reports/figures/forecasting/monthly_forecast.png", width = 8, height=4)

# Plot 2: The Components Breakdown
comp_plots <- prophet_plot_components(m_monthly, forecast_monthly)

wrap_plots(comp_plots, ncol=1)

ggsave("reports/figures/forecasting/monthly_components.png", width=8, height=6)
# plot 3: zoom in on holiday year-over-year

yoy_holiday_data <- forecast_daily %>%
  select(ds, holidays) %>%
  mutate(
    # Extract the actual year to use as our color category
    Year = as.factor(year(ds)),
    
    # Force every date to occur in 2024 (a leap year to protect Feb 29)
    # This aligns 2022, 2023, and 2024 exactly on top of each other
    dummy_date = update(ds, year = 2023)
  )

ggplot(yoy_holiday_data, aes(x = dummy_date, y = holidays, color = Year)) +
  # Baseline zero line
  geom_hline(yintercept = 0, color = "darkgray", linetype = "dashed") +
  
  # Draw the lines (alpha = 0.7 makes overlapping years easier to see through)
  geom_line(linewidth = .5, alpha = 0.7) +
  
  # Add dots only on the actual holidays
  geom_point(data = yoy_holiday_data %>% filter(holidays != 0), size = 1) +
  
  # Force the X-axis to label every single month using the dummy dates
  scale_x_datetime(date_breaks = "1 month", date_labels = "%b") +
  
  labs(
    title = "Year-Over-Year Holiday Effect on Cohort Quality",
    subtitle = "Baseline success rate multiplier during US National Holidays",
    x = "Month",
    y = "Holiday Effect",
    color = "Cohort Year"
  ) +
  theme_minimal()
```
