---
title: 'RNN / Sequential Processing'
subtitle: 'Stats M148 FingerHut Project - Second Round'
author: 'Charlie Kopp'
date: today
date-format: long
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
    fig-align: center
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(arrow)
library(data.table)
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)
```

## Introduction

In our previous steps, we flattened user journeys into static feature vectors for modeling with Random Forest and XGBoost. This notebook implements **Step 1: Sequential Data Preparation** for sequence models (like LSTMs). 

To optimize performance and maintain flexibility, we use **Apache Arrow (Parquet)**:
1. **Sequential Indexing**: We map `event_name` to dense sequential integers (`1` to `N`), reserving `0` for padding.
2. **Chronological Truncation**: We keep the last $M = 100$ events per user to capture the most recent context.
3. **Single-Cutpoint Sampling**: To keep memory usage manageable, we sample one random cutpoint per complete user journey for training.
4. **Cross-Platform Compatibility**: Parquet files can be read instantly in both R and Python.

---

## Load Data and Build Vocabulary Mapping

```{r}
#| label: load-raw-data

# Load datasets
dt <- setDT(read_parquet("data/processed/dt_clean2.parquet"))
user_outcomes <- setDT(read_parquet("data/processed/user_outcomes2.parquet"))

# Separate complete and incomplete user IDs
test_ids <- user_outcomes[final_outcome == "incomplete", id]
train_ids <- user_outcomes[final_outcome != "incomplete", id]

# Build event vocabulary mapping
unique_events <- sort(unique(dt$event_name))
vocab_mapping <- data.table(
  event_name = unique_events, 
  event_idx = seq_along(unique_events)
)

write_parquet(vocab_mapping, "data/processed/vocab_mapping.parquet")
```

---

## Sequential Data Preparation Function

```{r}
#| label: sequence-prep-function

prepare_sequential_data <- function(events_dt, vocab_map, max_len = 100) {
  setorder(events_dt, id, event_timestamp)
  
  # Calculate log-gaps
  events_dt[, ts_numeric := as.numeric(event_timestamp)]
  events_dt[, gap_s := ts_numeric - shift(ts_numeric), by = id]
  events_dt[is.na(gap_s), gap_s := 0]
  events_dt[, log_gap := log1p(gap_s)]
  
  # Map events
  events_dt <- merge(events_dt, vocab_map, by = "event_name", all.x = TRUE)
  
  # Truncate
  events_dt[, seq_rev_idx := rev(seq_len(.N)), by = id]
  events_truncated <- events_dt[seq_rev_idx <= max_len]
  
  setorder(events_truncated, id, event_timestamp)
  events_truncated[, seq_idx := seq_len(.N), by = id]
  
  output_dt <- events_truncated[, .(id, seq_idx, event_idx = event_idx, log_gap)]
  
  # Clean up memory
  events_dt[, c("ts_numeric", "gap_s", "log_gap", "event_idx", "seq_rev_idx") := NULL]
  
  return(output_dt)
}
```

---

## Single Cutpoint Training Snapshots

```{r}
#| label: single-cutpoint-sampling

# 1. Get journey bounds for training users
train_bounds <- dt[id %in% train_ids, .(
  start_ts = min(event_timestamp),
  end_ts   = max(event_timestamp)
), by = id]

# 2. Generate exactly one random cutpoint per user
set.seed(13)
train_bounds[, random_cutoff := start_ts + runif(.N) * as.numeric(difftime(end_ts, start_ts, units="secs"))]

# 3. Filter events to the cutpoint
dt[, event_ts_dup := event_timestamp]
train_events_subset <- dt[id %in% train_ids][
  train_bounds[, .(id, random_cutoff)], 
  on = .(id, event_ts_dup <= random_cutoff), 
  nomatch = NULL
]
setnames(train_events_subset, "event_ts_dup", "cutoff_time")

# Process Train Sequences
print("Processing training sequences...")
train_seq_dt <- prepare_sequential_data(train_events_subset, vocab_mapping, max_len = 100)

# Merge labels
train_seq_dt <- merge(
  train_seq_dt, 
  user_outcomes[, .(id, final_outcome)], 
  by = "id"
)

write_parquet(train_seq_dt, "data/processed/train_lstm_sequences.parquet")
rm(train_events_subset, train_seq_dt); gc()

# Test Sequences
print("Processing test sequences...")
test_events <- dt[id %in% test_ids]
test_events[, cutoff_time := max(event_timestamp)]
test_seq_dt <- prepare_sequential_data(test_events, vocab_mapping, max_len = 100)

write_parquet(test_seq_dt, "data/processed/test_lstm_sequences.parquet")
rm(test_events, test_seq_dt); gc()
```
