---
title: Assignment 1
subtitle: STATS M148 FingerHut Project
author: Charlie Kopp, Isaiah C. Mireles, Joel Yoon
date: today
date-format: long
format:
  pdf:
    toc: false
    number-sections: false
    geometry:
        - top=20mm
        - left=20mm
        - right=20mm
        - heightrounded
    execute:
      echo: true
      warning: false
      message: false
      include: false
    fig-align: center
    fig-height: 6
    fig-width: 8
---

```{r}
#| label: load packages
#| include: false
library(tidyverse)
```


# Task 0 - Fingerhut

[Generated GPT](https://chatgpt.com/g/g-69cee2d550f4819195683c49639b978c-fingerhut-gpt)

[Course Notes](https://docs.google.com/document/d/1k-IZ0ObDrfif1OiEVqaGXPZ5CbQxXZzn_GQ0rBYsBRE/edit?usp=sharing)

# Task 1 - Read data

```{r}
#| label: load data
data <- read_csv("C:/Users/charl/Desktop/UCLA/STATS M148/m148-project/data/dat_train1.csv")
```

```{r}
#| label: basic summary stats

nrow(data)
data |> distinct(id) |> nrow() # num unique ids

# earliest + latest timestamps
min(data$event_timestamp)
max(data$event_timestamp)
```

-   notice `id` feature is just **`customer_id account_id`**

# Task 2 - Remove Duplicates

- duplicates are when the same user logs the same action at the same time

```{r}
#| label: remove duplicates

# get total count of duplicate rows
data_unique <- data |> distinct(id, event_timestamp, event_name, .keep_all = TRUE)
n_dupes <- nrow(data) - nrow(data_unique)
n_dupes

n_dupes / nrow(data) # proportion of total

# count of clean rows
nrow(data_unique)

# actually delete duplicates
data <- data_unique
rm(data_unique)
```

# Task 3 - descriptive statistics 

1. How long are typical journeys? Actual time and actions.
2. What are common actions? Link the action IDs with the actual action
descriptions to answer this question.
3. How much time is between an individual user’s actions?
4. etc.


```{r}
#| label: data manipulation

# sample 10% of unique IDs to reduce memory usage
set.seed(133)
sampled_ids <- data |> distinct(id) |> slice_sample(prop = 0.1) |> pull(id)
data <- data |> filter(id %in% sampled_ids)

# 1. Journey lengths: time (in minutes) and total actions
journey_summaries <- data |>
	group_by(id) |>
	summarize(
		journey_duration_mins = as.numeric(difftime(max(event_timestamp), min(event_timestamp), units = "mins")),
		total_actions = n(),
		.groups = "drop"
	)

journey_stats <- journey_summaries |>
	summarize(
		mean_duration_mins = mean(journey_duration_mins, na.rm = TRUE),
		median_duration_mins = median(journey_duration_mins, na.rm = TRUE),
		mean_actions = mean(total_actions, na.rm = TRUE),
		median_actions = median(total_actions, na.rm = TRUE)
	)

# 2. Most common actions
common_actions <- data |>
	count(event_name, ed_id, name = "frequency") |>
	arrange(desc(frequency)) |>
	slice_head(n = 10)

# 3. Time between an individual user's actions
time_between <- data |>
	group_by(id) |>
	arrange(event_timestamp, .by_group = TRUE) |>
	mutate(
		time_since_last_action = as.numeric(difftime(event_timestamp, lag(event_timestamp), units = "mins"))
	) |>
	ungroup() |>
	filter(!is.na(time_since_last_action))

time_between_stats <- time_between |>
	summarize(
		mean_time_btw_actions = mean(time_since_last_action, na.rm = TRUE),
		median_time_btw_actions = median(time_since_last_action, na.rm = TRUE)
	)

# Bundle the summary objects into a single .RData file
save(
	journey_stats,
	common_actions,
	time_between_stats,
	file = "C:/Users/charl/Desktop/UCLA/STATS M148/m148-project/data/sample_summary_stats.RData"
)
```

# Task 4 - Flatten data (feature engineering)