---
title: "Sampling New Data"
author: "Isaiah C. Mireles"
date: "2026-05-26"
output: html_document
---

```{r}
dt <- fread("dat_train2.csv")
```

# Init. inspection and clean

```{r}
#| label: init inspection

# checking duplicates

sum(duplicated(dt, by=c("id", "event_timestamp", "event_name"), fromLast = TRUE))
sum(duplicated(dt, by=c("id", "event_timestamp", "event_name", "journey_steps_until_end"), fromLast = TRUE))

# there are (~3mil) duplicate events taken by same user at same time, but they are registered as separate actions (i.e. "double clicks")

# let's investigate:

# ensure temporal order
setorder(dt, id, event_timestamp)

idx_duped <- duplicated(dt, by=c("id", "event_timestamp", "event_name"), fromLast = TRUE)

# 1. Create a reference table of the duplicate keys
dupe_keys <- dt[idx_duped, .(id, event_timestamp, event_name)]

# 2. Use a join to pull all records matching those keys
verification_sample <- dt[dupe_keys, on = .(id, event_timestamp, event_name)]

# 3. View the first few sets
head(verification_sample[order(id, event_timestamp)], 10)


# Missing value summary
na_summary <- dt[, lapply(.SD, \(x) sum(is.na(x)))] |>
  melt(
    measure.vars = names(dt),
    variable.name = "column",
    value.name = "na_count"
  )
na_summary[, pct_missing := round((na_count / nrow(dt)) * 100, 2)]
na_summary

# no missing values!

# drop duplicates:
dt <- unique(dt, by=c("id", "event_timestamp", "event_name"))

# drop redundant / unused columns
dt <- dt[,-c("customer_id", "account_id", "journey_steps_until_end", "sep")]
```

# create user outcome dt

```{r}
#| label: create user outcome dt
train_cutoff_ts <- max(dt$event_timestamp)

user_outcomes <- dt[,
  .(
    has_shipped = any(ed_id == 28),
    last_ts = max(event_timestamp)
  ),
  by = id
]

user_outcomes[,
  days_inactive := as.integer(difftime(
    train_cutoff_ts,
    last_ts,
    units = "days"
  ))
]
user_outcomes[,
  final_outcome := fcase(
    has_shipped == TRUE , "success" ,
    days_inactive >= 60  , "failure" ,
    default = "incomplete"
  )
]

table(user_outcomes$final_outcome)

# Keep only the essential mapping
user_outcomes <- user_outcomes[, .(id, final_outcome)]
```

# Save cleaned dat

```{r}
#| label: save data

write_parquet(dt, 'dt_clean2.parquet')
write_parquet(user_outcomes, "user_outcomes2.parquet")
```

# Separate Complete and Incomplete (test and train) Users

```{r}
#| label: separate complete and incomplete users

# test_ids <- user_outcomes[final_outcome == "incomplete", id]
test_ids <- open_template$id
# train_ids <- user_outcomes[final_outcome != "incomplete", id]
train_ids <- setdiff(user_outcomes$id, test_ids)
```

# Define flattening function

```{r}
#| label: flatten function
flatten_journey <- function(events_dt) {
	setorder(events_dt, id, event_timestamp)

	# OPTIMIZATION 2: Create temporary integer columns for fast math.
	# This leaves original event_timestamp completely untouched for later use
	events_dt[, ts_int := as.integer(event_timestamp)]
	events_dt[, cutoff_int := as.integer(cutoff_time)]

	# OPTIMIZATION 3: Vectorize the gaps globally.
	# Calculating shift() inside a grouped block is slow. Doing it here is near-instant.
	events_dt[, gap_s := ts_int - shift(ts_int), by = id]

	# 1. Grouped Feature Engineering
	summary_dt <- events_dt[,
		{
			first_ts <- event_timestamp[1]
			last_ts <- event_timestamp[.N]

			first_ts_int <- ts_int[1]
			last_ts_int <- ts_int[.N]
			user_cutoff_int <- cutoff_int[1]

			.(
				first_action_ts = first_ts,
				last_action_ts = last_ts,

				# MATH HACK 1: Seconds length is just max minus min
				journey_length_s = last_ts_int - first_ts_int,

				# MATH HACK 2: Days is just the seconds difference divided by 86,400
				days_inactive = as.integer((user_cutoff_int - last_ts_int) / 86400),

				total_actions = .N,

				mean_gap_sec = if (.N > 1) mean(gap_s, na.rm = TRUE) else NA_real_,
				median_gap_sec = if (.N > 1) fmedian(gap_s, na.rm = TRUE) else NA_real_,
				max_gap_sec = if (.N > 1) max(gap_s, na.rm = TRUE) else NA_integer_
			)
		},
		by = id
	]

  # CLEANUP: Remove the temporary math columns
  events_dt[, c("ts_int", "cutoff_int", "gap_s") := NULL]

	# 2. Wide Event Counts
	counts <- dcast(
		events_dt,
		id ~ event_name,
		value.var = "event_name",
		fun.aggregate = length,
		fill = 0
	)
	setnames(
		counts,
		setdiff(names(counts), "id"),
		paste0("n_", setdiff(names(counts), "id"))
	)

	# 3. Merge and return
	return(merge(summary_dt, counts, by = "id"))
}
```

# Flatten test set

```{r}
#| label: flatten test set

test_events <- dt[id %in% test_ids]
test_events[, cutoff_time := max(event_timestamp)]
test_flat <- flatten_journey(test_events)
test_flat[, final_outcome := "incomplete"]

write_parquet(test_flat, "test_features2.parquet")
rm(test_events, test_flat); gc()
```

# Sample cut-points from train set

## 1. Simple Sample

-   single random cutpoint from each complete journey

```{r}
#| label: simple single sample
# 1. Filter raw events for Train IDs
train_events <- dt[id %in% train_ids]

# 2. Generate exactly one random time cutoff per user
set.seed(42)
cutoffs <- train_events[,
  .(
    start_ts = min(event_timestamp),
    end_ts = max(event_timestamp)
  ),
  by = id
]

# Generate a random timestamp between start and end
cutoffs[,
  random_cutoff := start_ts +
    runif(.N) * as.numeric(difftime(end_ts, start_ts, units = "secs"))
]

# 3. Join the cutoffs back to the events and filter

train_events <- train_events[cutoffs[, .(id, random_cutoff)], on = "id"]
train_events_truncated <- train_events[event_timestamp <= random_cutoff]
setnames(train_events_truncated, "random_cutoff", "cutoff_time")

# 4. Flatten the truncated training data
# Notice we pass 'random_cutoff' so days_inactive is calculated relative to the snapshot time
train_flat <- flatten_journey(train_events_truncated)

# 5. Bring in the true labels from step 1
# We must use the TRUE final outcome we calculated earlier, not the truncated outcome!

train_flat <- merge(train_flat, user_outcomes[id %in% train_ids], by = "id")

saveRDS(train_flat, "train_features_simple_sample2.rds")

rm(train_events, train_events_truncated, train_flat, cutoffs); gc()

# train_feats <- readRDS("data/train_features_simple_sample.rds")

# hist((train_feats$journey_length_s))

# rm(train_feats); gc()
```

## 2. Sample proportional to total_actions

```{r}
#| label: generate K cutoffs

train_events <- dt[id %in% train_ids]

set.seed(42)

# 1. Summarize user journeys to determine K
snapshot_counts <- train_events[, .(
  total_actions = .N,
  start_ts      = event_timestamp[1],
  end_ts        = event_timestamp[.N]
), by = id]

snapshot_counts <- merge(snapshot_counts, user_outcomes[, .(id, final_outcome)], by = "id")

# 2. Define K: bounded between 1 and 5

# frank(..., ties.method = "min") safely handles the heavy skew
# It ranks users 1 to N, divides by N to get a percentile (0 to 1), multiplies by 5, and rounds up.
snapshot_counts[, K := ceiling(5 * (frank(total_actions, ties.method = "min") / .N))]

# 3. Expand the table: Duplicate rows so each user has K rows
cutoffs <- snapshot_counts[rep(1:.N, K)]

# 4. Generate random cutoffs and a unique snapshot_id
cutoffs[, random_cutoff := start_ts + runif(.N) * as.numeric(difftime(end_ts, start_ts, units="secs"))]

# Create a unique ID for each snapshot (e.g., "User123_1", "User123_2")
cutoffs[, snapshot_id := paste0(id, "_", rowid(id))]
```

## 3. Sample proportional to journey_days

```{r}
#| label: generate K cutoffs per user based on journey_days

train_events <- dt[id %in% train_ids]

set.seed(13)

# 1. Summarize user journeys to determine K
snapshot_counts <- train_events[, .(
  total_actions = .N,
  start_ts      = event_timestamp[1],
  end_ts        = event_timestamp[.N]
), by = id][
  , journey_days := as.numeric(
    difftime(end_ts, start_ts, units="days"))]

snapshot_counts <- merge(snapshot_counts, user_outcomes[, .(id, final_outcome)], by = "id")

# 2. Define K: bounded between 1 and 5

# frank(..., ties.method = "min") safely handles the heavy skew
# It ranks users 1 to N, divides by N to get a percentile (0 to 1), multiplies by 5, and rounds up.
snapshot_counts[, K := ceiling(5 * (frank(journey_days, ties.method = "min") / .N))]

# 3. Expand the table: Duplicate rows so each user has K rows
cutoffs <- snapshot_counts[rep(1:.N, K)]

# 4. Generate random cutoffs and a unique snapshot_id
cutoffs[, random_cutoff := start_ts + runif(.N) * as.numeric(difftime(end_ts, start_ts, units="secs"))]

# Create a unique ID for each snapshot (e.g., "User123_1", "User123_2")
cutoffs[, snapshot_id := paste0(id, "_", rowid(id))]
```

### join K cutoffs back to train_events

```{r}
#| label: join K cutoffs back to train_events

# Duplicate the timestamp so it survives the join
train_events[, event_ts_dup := event_timestamp]

# Perform the Time Machine Join: Only keep events BEFORE the random cutoff
train_events_expanded <- train_events[cutoffs[, .(id, snapshot_id, random_cutoff)], 
                                      on = .(id, event_ts_dup <= random_cutoff), 
                                      nomatch = NULL, 
                                      allow.cartesian = TRUE]

# Cleanup column names post-join
setnames(train_events_expanded, "event_ts_dup", "cutoff_time") # The join condition column holds the cutoff timestamp

# Rename snapshot_id to 'id' so your function treats each snapshot as a unique user
setnames(train_events_expanded, "id", "original_user_id")
setnames(train_events_expanded, "snapshot_id", "id")


# Free up memory before flattening
rm(train_events, cutoffs, snapshot_counts); gc()


# Execute the flattening (This might take a few minutes depending on CPU)
train_flat <- flatten_journey(train_events_expanded)



# Rename 'id' back to 'snapshot_id' for clarity
setnames(train_flat, "id", "snapshot_id")

# extract the original ID out of the snapshot string
train_flat[, id := sub("_[0-9]+$", "", snapshot_id)]

# Add the final labels back to the flattened training data
train_flat <- merge(train_flat, 
                    user_outcomes[id %in% train_ids, .(id, final_outcome)], 
                    by = "id")

# Free up the expanded raw events
rm(train_events_expanded); gc()

# save k-cutoff sample
write_parquet(train_flat, "train_features_kcut_sample2.parquet")
```

# Modeling

```{r}
#| label: setup
#| include: false
library(data.table)
library(arrow)
library(mlr3)
library(mlr3tuning)
library(mlr3learners)
library(future)
library(xgboost)
```

# Prep Data

```{r}
#| label: load data

# ed_defs <- fread('data/Event Definitions.csv')
train_flat <- setDT(read_parquet("train_features_kcut_sample2.parquet"))
test_flat <- setDT(read_parquet("test_features2.parquet"))

train_flat[, final_outcome := factor(final_outcome, levels = c("success", "failure"))]
test_flat[, final_outcome := factor(final_outcome, levels = c("success", "failure"))]
```

## drop successes to reach 5%, keeping failures

```{r}
#| label: drop successes

# 1. Physical Downsampling to 5% (1:19 ratio)
successes <- train_flat[final_outcome == "success"]
failures  <- train_flat[final_outcome == "failure"]

# Failures / 19 = Number of successes needed to hit 5%
n_target <- floor(nrow(failures) / 19)

set.seed(42)
train_flat_balanced <- rbind(
  failures, 
  successes[sample(.N, n_target)]
)

# Free up the heavy temporary tables
rm(successes, failures, train_flat); gc()
```

## take toy sample (10% of train) with strat. sampling on outcome

```{r}
#| label: take toy sample

train_dev <- train_flat_balanced[, .SD[sample(.N, max(1, .N * 0.10))], by = final_outcome]
```

# mlr3 setup

## define classif task

```{r}
#| label: mlr3 setup

# Setting positive = "success" is important for precision/recall metrics later
tsk_train <- as_task_classif(
  train_flat_balanced, 
  target = "final_outcome", 
  positive = "success",
  id = "fh_model"
)

# Tell mlr3 that 'id' is a grouping factor, NOT a predictive feature
tsk_train$set_col_roles("id", roles = "group")
tsk_train$set_col_roles("snapshot_id", roles = "name")

# Tell mlr3 that 'snapshot_id' is just a row name, NOT a predictive feature

# drop leaky features
leakers <- c(
  'first_action_ts',
  'last_action_ts',
  'n_order_shipped',
  'n_place_order_web',
  "n_place_order_phone",
  "n_place_downpayment",
  "n_account_downpaymentreceived",
  "n_account_downpaymentcleared",
  'n_customer_requested_catalog_(digital)'
)

leakers <- intersect(leakers, tsk_train$feature_names)

tsk_train$set_col_roles(
  leakers,
  remove_from = "feature"
)

# make subsample task

tsk_dev <- as_task_classif(train_dev, target = "final_outcome", positive = "success")
tsk_dev$set_col_roles("id", roles = "group")
tsk_dev$set_col_roles("snapshot_id", roles = "name")
tsk_dev$set_col_roles(leakers, remove_from = "feature")
```

## define learners and rsmp

```{r}
#| label: learners and rsmp
lrn_rf <- lrn(
	'classif.ranger',
	id = 'randomForest_baseline',
	predict_type = 'prob',
	importance = 'impurity',
	num.trees = 500
)

lrn_xgb <- lrn(
	"classif.xgboost",
	id = "xgboost_tuned",
	predict_type = "prob",
	nthread = 8,
	booster = 'gbtree',
	tree_method = 'hist'
)

search_space <- ps(
	nrounds = p_int(lower = 50, upper = 300),
	eta = p_dbl(lower = 0.05, upper = 0.3),
	max_depth = p_int(lower = 3, upper = 8)
)

at_xgb <- auto_tuner(
	tuner = tnr("random_search", batch_size = 10),
	learner = lrn_xgb,
	resampling = rsmp("holdout"),
	measure = msr("classif.bbrier"),
	search_space = search_space,
	term_evals = 15 # 15 combinations is "light" but effective
)

# Standard Random Forest
lrn_rf_base <- lrn(
	"classif.ranger",
	id = "RF_Baseline",
	predict_type = "prob",
	importance = "impurity",
	num.trees = 100,
	max.depth = 15,
	num.threads = 8
)

# Standard XGBoost (No tuning)
lrn_xgb_base <- lrn(
	"classif.xgboost",
	id = "XGB_Baseline",
	predict_type = "prob",
	nrounds = 100,
	eta = 0.1,
	max_depth = 6,
	booster = 'gbtree',
	tree_method = 'hist',
	nthread = 8
)
```

## base models

```{r}
#| label: base models

rsmp_cv5 <- rsmp("cv", folds = 5)

measures <- list(msr("classif.bbrier"), msr("classif.prauc"))
```

-   pred didnt work

```{r}
rr_xgb_dev <- resample(tsk_dev, lrn_xgb_base, rsmp_cv5)
rr_xgb_dev$aggregate(measures)

rr_rf_dev <- resample(tsk_dev, lrn_rf_base, rsmp_cv5)
rr_rf_dev$aggregate(measures)

lrn_xgb_base$train(tsk_dev)

missing_cols <- setdiff(tsk_dev$feature_names, names(test_flat))

test_flat[, (missing_cols) := 0]


preds <- lrn_xgb_base$predict_newdata(test_flat)

submission <- data.table(
  id = test_flat$id,
  order_shipped = preds$prob[, "success"]
)


```

## tuning + bm

```{r}
plan(sequential)
options(future.globals.maxSize = 2 * 1024^3)



rsmp_cv5$instantiate(tsk_train)

# rr_rf <- resample(tsk_train, lrn_rf_base, rsmp_cv5)
rr_xgb <- resample(tsk_train, lrn_xgb_base, rsmp_cv5)
rr_xgb$aggregate(measures)

# TUNING WATCH OUT

# rr_xgb_tuned <- resample(
#   task = tsk_dev, 
#   learner = at_xgb, 
#   resampling = rsmp("cv", folds = 5) # Your standard grouped CV
# )

# rr_xgb_tuned$aggregate(measures)

at_xgb$train(tsk_train)
print(at_xgb$tuning_result)

preds <- at_xgb$predict_newdata(test_flat)

submission <- data.table(
  id = test_flat$id,
  order_shipped = preds$prob[, "success"]
)

# WAY too big for bmr...

# design <- benchmark_grid(
#   tasks = tsk_train,
#   learners = list(lrn_rf, at_xgb),
#   resamplings = rsmp("cv", folds = 5)
# )

# # Execute the benchmark
# bmr <- benchmark(design)
```

```{r}
library(vip)

# Train model first
lrn_rf_base$train(tsk_dev)

# Extract underlying ranger model
vip(lrn_rf_base$model) +
  ggplot2::labs(
    title = "Random Forest Variable Importance",
    x = "Importance",
    y = NULL
  )

lrn_rf_base$model$variable.importance
```

```{r}
library(vip)

lrn_xgb_base$train(tsk_dev)

vip(lrn_xgb_base$model) +
  ggplot2::labs(
    title = "XGBoost Variable Importance",
    x = "Importance",
    y = NULL
  )

xgb.importance(model = lrn_xgb_base$model)
```
